Statistical framework for whole-genome bisulfite-sequencing data

### 2014-12-08 Rmarkdown/wgbs_statistical_framework.Rmd
summary of wgbs_statistical_framework

There is a considerable amount of noise in the estimates of $\beta$-values when using low sequencing coverage, as can be seen from the low strand-correlations at these cutoffs. Once we require a minimum sequencing coverage of $5\times$, we see that most samples have a very high correlation of $\beta$-values across strands, $r = 0.8 - 0.9$. However, there are some notable exceptions. The embryonic stem cell samples --- _H1\_r1_, _H1\_r2_ and _HSF1_ from the Lister dataset; _J1\_1_ from the Seisenberger dataset_ --- all have far lower strand correlations, $r = 0.5 - 0.7$. This suggests that it may not be a good idea to combine CpG methylation levels across strands for embryonic stem cell samples. The other Seisenberger samples --- _E6.5\_epiblast\_1_ and _E16.5\_male\_1_ --- also have lower strand correlations, but these may be due to the fact that these data are from pooled samples.

Overall, it seems that non-embryonic stem cell samples have very high correlations of CpG $\beta$-values across strands, which means that these data can safely be combined across strands. However, it is a good idea to first check this assumption prior to combining data across strands.

@@

```{r Ziller_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
x <- read.table("../data/strand_correlation/Ziller_strand_correlations.txt",
stringsAsFactors = FALSE)
colnames(x) <- 1:10
xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the Ziller dataset. Each row is a sample and each column is a minimum coverage cutoff.", label = "tab:Ziller_strand_correlations")
align(xt) <- rep("c", ncol(x) + 1)
digits(xt) <- 2
print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE)
```

\subsection{Interpretation}

@@

In this chapter I have defined a mathematical framework for describing data from whole-genome bisulfite-sequencing data. I have sought to address some subtleties and complications that arise due to within-sample and between-sample differences in where DNA methylation is measured. Using this framework, I have described common estimators of DNA methylation and some of their statistical properties.

I have highlighted the genome-wide similarities and differences between a diverse set of whole-genome bisulfite-sequencing samples. I have shown that the CpG islands drive a strong bimodal distribution of $\beta$-values that is seen in almost all samples and that most intermediate methylation occurs outside of the CpG islands.

Somatic samples, particularly those from tissue rather than cell lines, have very 'well-behaved' $\beta$-value distributions that are globally similar between samples. The may reflect that the DNA methylome is well-established in these samples and relatively static. The increased level of partial methylation in somatic cell lines compared to somatic tissue samples may be attributable to the development of sub-clones during the culturing of the sample. This is consistent with the high epipolymorphism observed in a study that tracked the dynamics of DNA methylation in an _in vitro_ evolutionary cell culture system \citep{Landan:2012kp}. Somatic samples also have very highly correlated CpG $\beta$-values across strands, meaning that these data can generally be combined across strand to increase the sequencing coverage of each CpG.

Induced pluripotent stem cell lines also appear to have a quite strictly regulated methylome, with little intermediate methylation. This is likely a consequence of the fact that during the induction of pluripotency, the methylome of the sample is 'reset' \citep{Stricker:2013kl} thereby resulting in a homogeneous population of cells.

The most distinct methylomes are the hypermethylated embryonic stem cells (_H1\_r1\_ and _H1\_r2_), the hypervariable progenitor germ cells (_E16.5\_male\_1_) and the hypomethylated cancer cell lines (_SRR949206_ and _SRR949207_). __TODO: SOmethign about Seisenberger being difficult to interpret__

\section{General TODOs}

* Discuss distribution of coverage? Perhaps in context of sequencing bias?

>>>>

There is a considerable amount of noise in the estimates of $\beta$-values when using low sequencing coverage, as can be seen from the low strand-correlations at these lower coverage cutoffs. Once we require a minimum sequencing coverage of $5\times$, we see that most samples have a very high correlation of $\beta$-values across strands, $r = 0.8 - 0.9$, however, there are some notable exceptions.

The embryonic stem cell samples --- _H1\_r1_, _H1\_r2_ and _HSF1_ from the Lister dataset --- all have less correlated strand-specific $\beta$-values, $r = 0.5 - 0.7$. The other embryonic stem cell samples have higher correlations of $\beta$-values across strands, although the estimates are quite different between the two replicates of the same cell line (_H9_: $r = 0.77$, _H9\_Laurent_: $r = 0.92$). This suggests that it may not be a good idea to combine CpG methylation levels across strands for embryonic stem cell samples.

All three Seisenberger samples have noticeably less correlated strand-specific $\beta$-values, including the embryonic stem cell _J1\_1_. However, since these data are from pooled DNA, the source of this reduced correlation is difficult to pin down.

Overall, it seems that with the exception of embryonic stem cells, most samples have very highly correlated strand-specific CpG $\beta$-values, which means that these data can safely be combined across strands. However, it remains a good idea to first check this assumption prior to combining data across strands.

@@

```{r Ziller_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
x <- read.table("../data/strand_correlation/Ziller_strand_correlations.txt",
stringsAsFactors = FALSE)
colnames(x) <- 1:10
xt <- xtable(x, caption = "Correlations of $\\beta$-values across strands for the Ziller dataset. Each row is a sample and each column is a minimum coverage cutoff.", label = "tab:Ziller_strand_correlations")
align(xt) <- rep("c", ncol(x) + 1)
digits(xt) <- 2
print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE)
```
\subsection{Interpretation}

@@

In this chapter I have defined a mathematical framework for describing data from whole-genome bisulfite-sequencing data. I have sought to address some subtleties and complications that arise due to within-sample and between-sample differences in where DNA methylation is measured. By using a common statistical framework we can better understand how different statistical methods relate to one another. Using this framework, I have described common estimators of DNA methylation levels and some of their statistical properties. I then examined the empirical distributions of these variables across a diverse set of 51 whole-genome bisulfite-sequencing samples.

Using these data I have shown that the CpG islands drive the strong bimodal distribution of $\beta$-values that is seen in almost all samples and that most intermediate methylation occurs outside of the CpG islands. The most distinct methylomes are the hypermethylated embryonic stem cells (_H1\_r1\_ and _H1\_r2_) and the hypomethylated cancer cell lines (_SRR949206_ and _SRR949207_). The Seisenberger samples also stand out, particularly the hypervariable progenitor germ cells (_E16.5\_male\_1_), however, these samples are more difficult to interpret at the genome-wide level since they are from pooled DNA. Nonetheless, some of the increased variability in the Seisenberger data will also reflect that these samples are from developmental timepoints during which DNA methylation is very dynamic.

In contrast, the somatic samples, particularly those from tissue rather than cell lines, have very 'well-behaved' $\beta$-value distributions that are globally similar between samples. The may reflect that the DNA methylome is well-established in these samples and relatively static. The increased level of partial methylation in somatic cell lines compared to somatic tissue samples may be attributable to the development of sub-clones during the culturing of the sample. This is consistent with the high epipolymorphism observed in a study that tracked the dynamics of DNA methylation in an _in vitro_ evolutionary cell culture system \citep{Landan:2012kp}. Somatic samples also have very highly correlated CpG $\beta$-values across strands, meaning that these data can generally be combined across strands to increase the sequencing coverage of each CpG.

Induced pluripotent stem cell lines also appear to have a quite strictly regulated methylome, with little intermediate methylation. This is likely a consequence of the fact that during the induction of pluripotency, the methylome of the sample is 'reset' \citep{Stricker:2013kl} thereby resulting in a homogeneous population of cells. These samples also have highly correlated strand-specific $\beta$-values, meaning that these data can generally be combined across strands to increase the sequencing coverage of each CpG.

All of the above highlights the considerable variability of DNA methylation data, both between samples and within a sample, and the care with which $\beta$-values must be interpreted. While an attractively simple measure, analyses based on $\beta$-values are based on the 'average' behaviour, where the averaging is over many sources of variation. Analyses based on $\beta$-values also do not make full use of the information available in bisulfite-sequencing data, as we shall see in Chapter \ref{chap:wgbs_downstream_analyses}.
